From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-02-27 appendixB/appendixB.Rmd
WIP

A derivation of diffusion approximation is originally credited to @Kramers1940 and later improved by @Moyal1949, though Einstein makes implicit use of this approximation in his own work on Brownian Motion decades earlier [@Gardiner2009].  The diffusion approximation has been a popular basis for ecological stochastic models as well, such as in @Nisbet1982's classic textbook on fluctuating populations, or more recently in @Lande2003 and work reviewed in @Ovaskeinen.  This diffusion approximation is nicely summarized @Gardiner2009 and in @vanKampen2007, though I believe @Kurtz1978 provides a more formal proof that goes unmentioned by any of these references.  Both the diffusion approximation and the van Kampen expansion start with the same master equation that defines the underlying dynamics of births and deaths (matching the process implemented in the exact Gillespie algorithm, @Gillespie1977, see Appendix A).

>>>>

A derivation of diffusion approximation is originally credited to @Kramers1940 and later improved by @Moyal1949, though Einstein makes implicit use of this approximation in his own work on Brownian Motion decades earlier [@Gardiner2009].  The diffusion approximation has been a popular basis for ecological stochastic models as well, such as in @Nisbet1982's classic textbook on fluctuating populations, or more recently in @Lande2003 and work reviewed in @Ovaskainen2010.  This diffusion approximation is nicely summarized @Gardiner2009 and in @vanKampen2007, though I believe @Kurtz1978 provides a more formal proof that goes unmentioned by any of these references.  Both the diffusion approximation and the van Kampen expansion start with the same master equation that defines the underlying dynamics of births and deaths (matching the process implemented in the exact Gillespie algorithm, @Gillespie1977, see Appendix A).

### 2018-02-27 paper/paper.Rmd
WIP

As the literature on stochastic modeling in ecology is both extensive and often
technical, and so rather than attempt a comprehensive review, this synthesis
will focus on the simplest examples necessary to illustrate this broader trend
in the changing way we think about the role of noise and the opportunity
that shift can present to better bridge theoretical and empirical communities.
To that end, I will focus on examples from simple population models, echoing
the motivation of @Bartlett1960 to focus on basic theoretical aspects common
across models. In so doing, we will ignore the equally rich literature of 
stochastic models in evolution [@Lenormand2009] and population genetics
[e.g. @Ewens2004]. In an effort to make these themes accessible to more emprical
readers, I will also focus on simplest examples necessary, with only passing
mention of the rich literature extending these concepts to more realistic
models.  It is easy to critique this choice: indeed, one might summarize decades
of research in noting the twin observations that (1)  ecology is complex and (2)
that complexity is important. In the stochasticity literature alone, much work 
has been done to understand and extend fundamental results to more complex models,
including the importance of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989;
@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g. @Durrett1994; @Dieckmann2000;
@Kerr2002; @Schreiber2010], Individual heterogeneity [e.g. @Vindenes2015; @Hart2016]
or the coexistence of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Schreiber2017]; 
correlated [@Ripa; @Ripa1999; @Petchey, @Schreiber2010; @Lee2017; @Spanio2017] or periodic
[@Coulson2001; @Bjornstad2001; @Keeling2001] structure in environmental noise, or
the interaction of ecological and evolutionary processes [@Saether1997; @Ozgul2009;
@Lenormand2009; @Vindenes2015; @Schreiber2015]. In this we follow @Levins1966 in
sacrificing realism for generality.

@@

where $\left[(1-\frac{\bar{e}}{c}) N$ is the equilibrium population size.
This reflects the assumption of the canonical equation, Eq \eqref{canonical}
that the standard deviation scales in direct proportion to the system size $N$,
while the demographic noise scales as the square root.  Consequently,
for a large system, the demographic contribution becomes proportionally smaller
while the environmental contribution remains the same.

>>>>

The gap between theory-focused specialists and more empirically minded non-specialists
is particularly acute in the ecological literature on stochastic processes, making
a review of

As the literature on stochastic modeling in ecology is both extensive and often
technical, and so rather than attempt a comprehensive review, this synthesis
will focus on the simplest examples necessary to illustrate this broader trend
in the changing way we think about the role of noise.
To that end, I will focus on examples from simple population models, echoing
the motivation of @Bartlett1960 to focus on basic theoretical aspects common
across models. In so doing, we will ignore the equally rich literature of 
stochastic models in evolution [@Lenormand2009] and population genetics
[e.g. @Ewens2004]. In an effort to make these themes accessible to more emprical
readers, I will rely on the simplest examples necessary, with only passing
mention of the rich literature extending these concepts to more realistic
models.  It is easy to critique this choice: indeed, one might summarize decades
of research in noting the twin observations that (1)  ecology is complex and (2)
that complexity is important. In the stochasticity literature alone, much work 
has been done to understand and extend fundamental results to more complex models,
including the importance of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989;
@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g. @Durrett1994; @Dieckmann2000;
@Kerr2002; @Schreiber2010], Individual heterogeneity [e.g. @Vindenes2015; @Hart2016]
or the coexistence of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Schreiber2017]; 
correlated [@Ripa; @Ripa1999; @Petchey, @Schreiber2010; @Lee2017; @Spanio2017] or periodic
[@Coulson2001; @Bjornstad2001; @Keeling2001] structure in environmental noise, or
the interaction of ecological and evolutionary processes [@Saether1997; @Ozgul2009;
@Lenormand2009; @Vindenes2015; @Schreiber2015]. In this we follow @Levins1966 in
sacrificing realism for generality.

@@

where $(1-\frac{\bar{e}}{c}) N$ is the equilibrium population size.
This reflects the assumption of the canonical equation, Eq \eqref{canonical}
that the standard deviation scales in direct proportion to the system size $N$,
while the demographic noise scales as the square root.  Consequently,
for a large system, the demographic contribution becomes proportionally smaller
while the environmental contribution remains the same.
